\documentclass[11pt,a4paper]{letter}
\usepackage[margin=2.5cm]{geometry}
\usepackage{times}
\usepackage{hyperref}

\signature{Jin Ho Choi, Member, IEEE\\
SmartEAR, Seoul, South Korea\\
\texttt{jinhochoi@smartear.co.kr}}

\address{Jin Ho Choi\\
SmartEAR\\
Seoul, South Korea\\
\texttt{jinhochoi@smartear.co.kr}}

\begin{document}

\begin{letter}{Editor-in-Chief\\
IEEE Signal Processing Letters}

\opening{Dear Editor,}

I am submitting the manuscript entitled:

\begin{center}
\textbf{``When Does Temporal Memory Help? Selective SSM vs.\ MLP\\
in RL-Based Multi-Target DOA Track Management''}
\end{center}

for consideration for publication in \textit{IEEE Signal Processing
Letters} as an original Letter.

\textbf{Summary.}
This letter addresses a fundamental question in reinforcement
learning (RL)-based adaptive tracking: when does adding temporal
memory to an RL policy improve performance over a memoryless
multilayer perceptron (MLP)?
We prove, via the data processing inequality and the concept of a
sufficient statistic, that a selective state space model (SSM) encoder
outperforms an MLP if and only if the observation is not a sufficient
statistic for the policy-relevant history.
We instantiate this result as \textbf{Mamba-COP-RL}, a novel
architecture coupling a Mamba SSM encoder with a PPO actor-critic
operating on raw 181-dimensional COP spatial spectra.
Experiments on \textbf{CombatTrackEnv}, an open-source benchmark
with four combat-realistic multi-target DOA tracking scenarios,
demonstrate GOSPA improvements of up to 25.2\% over fixed GM-PHD
thresholds and 6.7\% over MLP-PPO, with gains ordered by temporal
dependency horizon as predicted by the theory.
Ablation against GRU-PPO and LSTM-PPO confirms that Mamba's
input-dependent selectivity is critical.
The 41.4\,KB INT8 model achieves 3.4\,ms/scan latency on an
STM32H7 Cortex-M7, enabling edge deployment.

\textbf{Revisions.}
The manuscript has been carefully revised to improve clarity and
conciseness.
Specifically: (i) the abstract was condensed to highlight the
core theoretical and empirical contributions;
(ii) the related work section was restructured to emphasize
Mamba's intrinsic selective-gating motivation rather than
downstream application domains;
(iii) Algorithm~1 was removed and replaced with a concise BPTT
description to reduce length;
(iv) all references were verified against IEEE Xplore, arXiv,
and DBLP, with author lists, titles, venues, and DOIs corrected
where necessary.

\textbf{Scope.}
This work is squarely within the scope of IEEE Signal Processing
Letters: it introduces a signal processing architecture (Mamba SSM
encoder on COP spatial spectra), provides a formal theoretical
proposition on sequential signal processing, and demonstrates
edge deployment on resource-constrained hardware.

\textbf{Related submissions.}
The corresponding author has a related manuscript currently submitted
to \textit{IEEE Transactions on Signal Processing}:
``Underdetermined High-Resolution DOA Estimation and Multi-Target
Tracking via COP-RFS'' (J.\,H.\ Choi, 2026, cited as Ref.~[3]).
That work introduces the COP-RFS tracking framework; the present
Letter focuses exclusively on the RL policy architecture and
formally characterizes when temporal memory is beneficial.
The two contributions are complementary and non-overlapping.

\textbf{Declarations.}
This manuscript has not been published previously and is not under
review at any other journal or conference.
No preprint has been posted on any server.

\closing{Sincerely,}

\end{letter}
\end{document}
